A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-11-06 imaging_demo/ms.tex
Update section 3 with suggested changes
From 3b460d9 (review):
=====================

In ms.tex,

(*) Update section 3 with suggested changes

EPIC can accept dual-polarization inputs and produce images of all four 
cross-polarizations. Currently two data input formats exist for reading in the 
electric field time samples measured by the antennas -- simulated electric 
fields based on a sky model using the simulator packaged with EPIC; and LWA 
data. Efforts to build interfaces for data from other telescopes are underway.

@@

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{ij}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are theoretically 
identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $V_\textrm{ij}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Antenna-to-Grid Mapping}
\par\medskip
\noindent A grid is generated on the coordinate system in which antenna 
locations are specified with a grid spacing. The grid spacing can be controlled 
by the user. By default, it is set to be at $\lambda_\textrm{min}/2$ even at the 
highest frequency to ensure there is no aliasing even from regions of the sky 
far away from the field of view. The number of locations on the grid is 
restricted to be a power of 2.

@@

In the most general case, this gridding kernel could contain information on
$w$-projection effects, and other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$\Ngrid\times \Nant$, reduces to a sparse block-diagonal matrix
with only $\Nant$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. This sparse matrix is stored in a Compressed Sparse Row (CSR) format. 
Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following 
the MOFF algorithm places no restriction on array layout, and accounts for 
non-identical antenna beam patterns. Because only a single correlation is needed 
for each antenna, the computation complexity scales only as 
$\sim N_{\mathrm{a}}N_\textrm{c}$, where $N_\textrm{c}$ is the number of calibrators 
in the field.

The calibration module included in the EPIC repository allows for application of 
pre-determined calibration solutions, or can solve for the complex gains using 
the EPICal algorithm.

@@

\par\medskip
\noindent {\bf Spatial Fourier Transform}
\par\medskip
\noindent Before the spatial Fourier transform, the gridded electric fields are 
padded with zeros in order to match the grid size and angular size of each image 
pixel that would have been obtained with software holography of output from an FX
correlator.

@@

The other function (not shown in the figure) is to update the data as new data 
keeps streaming in. This can also be parallelized. Another important attribute
(not shown in figure) are antenna flags for each polarization appropriate for 
the data stream being held by the module.

@@

The interferometer module holds the attributes and functions pertaining to
a pair of antennas and represents the cross-correlation information obtained
from the pair. Its primary attributes are the two antenna modules. It also
contains four cross-polarized visibility time-series and spectra.

@@

Besides its core functions of spatial Fourier transform and squaring, it can 
stack, accumulate and average images, and optionally remove the antenna
auto-correlations centered around the zero-spacing pixel in the $uv$ plane. 
It also handles all four cross-polarization products. Currently, it can 
support writing data out in standard FITS format.

>>>>

EPIC can accept dual-polarization inputs and produce images of all four 
instrumental cross-polarizations. Currently two data input formats exist for 
reading in the 
electric field time samples measured by the antennas -- simulated electric 
fields based on a sky model using the simulator packaged with EPIC; and LWA 
data. Efforts to build interfaces for data from other telescopes are underway.

@@

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{p}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are theoretically 
identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $V_\textrm{p}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Antenna-to-Grid Mapping}
\par\medskip
\noindent A grid is generated on the coordinate system in which antenna 
locations are specified with a grid spacing. The grid spacing can be controlled 
by the user. By default, it is set to be at $\lambda/2$ even at the 
lowest wavelength to ensure there is no aliasing even from regions of the sky 
far away from the field of view. The number of locations on the grid is 
restricted to be a power of 2 for efficient use of FFT.

@@

In the most general case, this gridding kernel could contain information on
$w$-projection effects, and other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the 
antennas ($\mathbf{a}$) to the antenna array grid denoted by the coordinate 
$\mathbf{r}$. Since each antenna occupies a footprint typically the size of its 
aperture, $\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$\Ngrid\times \Nant$, reduces to a sparse block-diagonal matrix
with only $\Nant$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block, where $N_\textrm{k}$ is the number of grid points that fall inside an
antenna's footprint. This sparse matrix is stored in a Compressed Sparse Row 
(CSR) format. Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following 
the MOFF algorithm places no restriction on array layout, and accounts for 
non-identical antenna beam patterns. Direction dependent calibration can be 
achieved by correlating antenna signals with output pixels in the direction of 
$N_\textrm{c}$ calibration sources, then fitting for a functional model of the 
sky. Since antennas are only correlated with calibrator pixels, the 
computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. % And since 
% only a single correlation is needed for each antenna, the computation complexity 
% scales only as $\sim N_{\mathrm{a}}N_\textrm{c}$, where $N_\textrm{c}$ is the 
% number of calibrators in the field.

The calibration module included in EPIC allows for application of pre-determined 
calibration solutions, or can solve for the complex gains using the EPICal 
algorithm.

@@

\par\medskip
\noindent {\bf Spatial Fourier Transform}
\par\medskip
\noindent Before the spatial Fourier transform, the gridded electric fields are 
padded with zeros in order to match the grid size and angular size of each image 
pixel that would have been obtained with the software holography output from an 
FX correlator.

@@

The other function (not shown in the figure) is to update the data as new data 
streams in. This can also be parallelized. Another important attribute
(not shown in figure) are antenna flags for each polarization appropriate for 
the data stream being held by the module.

@@

The interferometer module holds the attributes and functions pertaining to
a pair of antennas and represents the cross-correlation information obtained
from the pair. Its primary attributes are the two antenna modules. It also
contains four cross-polarized visibility time-series (even for the FX 
correlator for diagnostic purposes) and spectra.

@@

Besides its core functions of spatial Fourier transform and squaring, it can 
stack, accumulate and average images, and optionally remove the antenna
auto-correlations centered around the zero-spacing pixel in the $uv$ plane. 
It also handles all four cross-polarization products. Currently, it supports 
writing data out in standard FITS format.
